A Constructive Ordered Predicate Logic for Category Theory

### 2022-10-13 paper/paper.tex
tighten up intro

\begin{abstract}
  We present a domain-specific type theory for constructions and
  proofs in category theory. The type theory axiomatizes notions of
  category, functor, profunctor and a generalized form of natural
  transformations. The type theory imposes an ordered linear
  restriction on standard predicate logic, which guarantees that all functions
  between categories are functorial, all relations are profunctorial,
  and all transformations are natural by construction, with no
  separate proofs necessary. Important category theoretic proofs such
  as the Yoneda lemma and Co-yoneda lemma become simple type theoretic
  proofs about the relationship between unit, tensor and (ordered)
  function types, and can be seen to be ordered refinements of
  theorems in predicate logic.

  We develop a sound and complete notion of categorical model for the
  type theory based on virtual equipments, which are known to model
  both internal and enriched category theory. This means that while
  the proofs in our type theory look like standard set-based
  arguments, the syntactic discipline ensure that all proofs and
  constructions carry over to enriched and internal settings as well.
\end{abstract}

@@

Category theory is a branch of mathematics that studies
higher-dimensional typed algebraic structures .
%
Originally developed for applications to homological algebra, it was
quickly discovered that (among many other applications) these
categorical structures were common in logic and computer science, and
formal systems like logics, type theories and programming languages
typically have sound and complete models given by notions of
structured categories\cite{lawvere,goguen,lambek,moggi}, a phenomenon
sometimes referred to as the Curry-Howard-Lambek nncorrespondence
\cite{curry-howard-lambek}.
%
Examples include the simply typed lambda calculus \cite{lambekscott},
computational lambda calculus \cite{moggi}, linear logic \cite{girard}
and dependent type theory\cite{cartmell1986}.
%
These seminal works have led to a fruitful field of research known as
categorical logic, where new type theories are designed based directly
on category-theoretic semantics.
%
The slogan is that the syntax of a type theory should present an
initial object in its category of models, a category-theoretic
reformulation and generalization of the logical notion of soundness
and completeness.

% TODO: do we have citations for this?
While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex, and there is difficulty in
acquisition of competence in category theory by computer scientists
who do not have a strong background in pure mathemantics.
%
While much of this can be explained by lack of, or lack of awareness
of, accessible material for computer scientists, in some ways category
theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and limits produce a proliferation of ``naturality'' and
``functoriality'' side-conditions that must be discharged.
%
This is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
These generalizations have long been recognized as being important in
programming language applications: domain-, metric- and
step-index-enriched categories have been used to model recursive
programming languages and internal categories have been used to model
parametricity and gradual
typing\cite{order-enriched-cats,topos-of-trees,reflexive-graphs,double-cats-gradual-typing}.

For instance, when constructing an adjoint pair of functors between
two categories, a na\"ive approach would define all of the data of the
action on objects, action on arrows, prove the functoriality of such
actions, as well as construct two families of transformations, prove
they are natural and then finally proving a pair of equalities
relating compositions of natural transformations.
%
Furthermore, when moving into the enriched or internal setting, the
proofs can have further continuity, monotonicity, or other
side-conditions.
%
Carrying out the proofs explicitly is quite tedious and many newcomers
to are left with the impression that category theory is full of long,
but ultimately trivial constructions.

Fortunately, the tools of category theory itself can be employed to
simplify these constructions, specifically the tools of \emph{higher}
category theory.
%
Consider the analogous situation in differential calculus. When an
adept analyst writes down a function, they do not expand out the
$\epsilon\mathord{-}\delta$ definition of continuity for a function
and proceed from first principles. Rather, they are aware of certain
\emph{syntactic principles} for defining functions that are continuous
by construction, most importantly that composition of continuous
functions is continuous and that previously defined functions are
known to be continuous.
%
Similar principles apply to category theory itself: functors and
natural transformations are closed under composition and whiskering
operations, and experienced category theorists rely on these syntactic
principles to eliminate the tedium of explicit proofs.
%
In the case of category theory, these principles can be formalized
using algebraic structures such as 2-categories, bicategories,
(virtual) double categories and pro-arrow equipments\cite{TODO}, an
approach known as \emph{formal category theory}.
%
In these structures, rather than defining notions of category, functor
and natural transformation from first principles, they are axiomatized
in a manner similar to how a category axiomatizes a notion of space
and homomorphism.
%
One benefit is that the formal approaches also apply to enriched and
internal settings, so \emph{formal} category theory proofs apply to
these settings.
%
A downside is that these algebraic structures are quite complicated,
and practitioners typically employ either an algebraic combinator
syntax or a 2-dimensional diagrammatic language that can be quite
beautiful and elegant, but is also seems very far removed from the
traditional formulation of category theory in terms of sets and
functions.

In this work, we apply the techniques of categorical logic to define a
more familiar logical syntax for carrying out constructions and proofs
in formal category theory.
%
We call the resulting theory \emph{virtual equipment type theory}
(VETT) as (hyperdoctrines of) virtual equipments provide a sound and
complete notion of model for the theory.
%
In VETT, functors and natural transformations are defined using
ordinary term syntax, using familiar notations and $\beta\eta$
reasoning principles for $\lambda$-functions, bound variables, tuples,
etc.
%
By adhering to a \emph{syntactic discipline}, the logic guarantees
that all functor terms are automatically functorial, all and natural
transformation terms are natural.
%
More specifically, the syntax for transformations is a kind of
\emph{indexed, ordered linear} $lambda$ calculus, where the indexing
ensures that transformations are correctly natural and the ordering
and linearity ensure that the proofs are valid in a large class of
enriched and internal categories: for instance enrichment in a
non-symmetric monoidal category.
%
VETT provides an alternative alternative to algebraic and
string-diagram syntaxes for working with virtual equipments,
similar to how the lambda calculus provides an alternative to
categorical combinators and string diagram calculi for cartesian
closed categories.

In this work, we apply the technique of categorical logic to these
algebraic structures to design a formal logic, which we call
\emph{virtual equipment type theory}, that can be soundly interpreted
in virtual equipments.
%
In the ``standard model'' of \vett{}, the sorts are interpreted as
categories, functions as functors, relations as profunctors and
implications of relations as profunctor homomorphisms, a generalization
of natural transformations.
%
The type theoretic syntax allows us to write all of our constructions
as if they were ordinary functors and profunctors in set-based
category theory.
%
\vett also can be modeled in an arbitrary virtual equipments, and so the
constructions are valid in enriched and internal category theory as
well.
%%
%% DRL: I know what you mean, but I think conflating set theory and
%% type theory will put some people off... 
%% 
%% This is analogous to the situation in intuitionistic type theory,
%% where the syntax is similar to that of informal set theory, but by
%% restricting to intuitionistic reasoning, the constructions can all be
%% interpreted in models where types are spaces and terms are continuous
%% functions ``for free''.
%

The syntax of \vett{} is a kind of indexed, ordered linear,
proof-relevant variant of predicate logic over a unary type theory.
%
That is just as a predicate logic has a notion of type, term,
relation and implication, \vett{} is based on four analogous category
theoretic concepts: categories, functors, profunctors and natural
transformations of profunctors.
%
So categories are simply treated like types, and functors are simply
terms whose type is a category and whose free variables have
categories as types as well. The functor syntax is a unary type
theory, meaning functor terms always have precisely one free variable
representing the input object.
%
The analogue of a relation is a \emph{profunctor} (defined below),
which are written as a kind of set with free category variables. In
keeping with our models, like the restriction to unary functors, we
restrict to profunctors with two free variables.
%
The logic is proof-relevant in that the implications of relations are
generalized to natural transformations of profunctors, and we use a
$\lambda$-calculus notation to describe these ``proof terms''.
%
We can strenghten this analogy into a formal construction: any
construction in \vett{} can be erased to a corresponding construction
or proof in predicate logic, as sets, functions and relations define a
(somewhat degenerate) virtual equipment.
%
Our syntax highlights this analogy in that some of our constructions
are generalizations of set comprehension or power-sets and we use
similar notation.

The restriction to unary type theory and binary relations is fairly
strict: we cannot discuss important concepts such as functor
categories or opposite categories. However, the restriction is natural
in that it corresponds to a well-understood notion of model (virtual
equipments).
%
It is well known that a great deal of fundamental results and
constructions in category theory, and they have been used for this
purpose in the formal category theory literature
\cite{elements-of-oo-category-theory, enriched-indexed-categories}.
%
However, we can work around these unary/binary restrictions to some
extent by viewing the type theory as a domain-specific language
embedded in a metalanguage.  For example, while we cannot talk about
functor categories inside the ordered part of the type theory, we can
state a theorem that quantifies over functors using the
meta-language's universal quantifier.  To more easily formalize these
metatheorems, we add a third layer to the type theory: an extensional
dependent type theory in the style of Martin-L\"of type theory. Then
all of our ordered predicate logic judgments will also be indexed by a
context from this dependent type theory, and the type theory includes
universe types for categories, functors, profunctors and natural
transformations that allow us to formalize theorems the logic is too
restrictive to encode, analogous to the use of 2-level type theory in
homotopy type theory \cite{voevodsky12,altenkirchXX}.

@@

While we emphasize the applications to enriched and internal category
theory in this work, there is potential for more direct application to
programming language semantics.
%
Ordinary predicate logic has been used to develop logics in which
logical relations proofs can be developed such as Abadi-Plotkin logic
for parametricity and LSLR and Iris for step-indexed logical relations
proofs \cite{abadi-plotkin,lslr,iris}.
%
We conjecture that \vett{} might similarly serve as the foundation for
a logic of \emph{ordered} structures, which abound in applications:
rewriting and approximation relations can both be modeled as orderings
and logical relations involving these structures are proven to respect
orderings: operational logical relations must be downward-closed and
approximation relations should satisfy transitivity.
%
Just as LSLR and Iris release the user from the syntactic burden of
explicit step-indexing, \vett{} may be used to release the user from
the syntactic burden of proving downward-closure or transitivity
side-conditions.
%
Additionally, \vett{} may serve as the basis of a future domain
specific proof assistant for category theoretic proofs.

\paragraph{Basics of Profunctors.}

While we assume the reader has some background knowledge of category
theory, we briefly define profunctors, which are not included in many
introductory texts.  Profunctors arise implicitly in formulations of
universal properties such as representable functors and adjunctions.

First, a category $\cat C$ has a collection of objects and morphisms
with identity and composition, and a functor $F : \cat C \to \cat D$ is
a function on objects and a function on morphisms that respects source
and target and preserves identity and composition. It will be useful for
intuition to think of a category as a generalization of a preordered
set, which has a set of elements and a binary \emph{relation} on its
objects satisfying reflexivity and transitivity. It is fruitful to think
of a category as a \emph{proof-relevant preorder}, where morphisms are
the proofs of ordering, and the reflexivity and transitivity
constructions must additionally satisfy equations that make them act
like identity and composition. Then a functor is a proof-relevant
version of a monotone function.

Given categories $\mathcal C$ and $\mathcal D$, a profunctor $R$ from
$\mathcal C$ to $\mathcal D$, written $R : \cat C \pto \cat D$ is a
functor $R : {\cat C}^{op} \times {\cat D} \to \Set$. We can think of
each set $R(c,d)$ as a set of ``heteromorphisms'' from $c$ (an object of
$\cat C$) to $d$ (an object of $\cat D$).  These ``heteromorphisms''
differ from the usual ``homomorphisms'' in a single category $\cat C$
because they relate two objects from two different categories.
Functoriality then defines a notion of composition of heteromorphisms
with homomorphisms in $\cat C$ and $\cat D$. Thinking of categories as
proof-relevant preorders, a profunctor is a proof-relevant relation that
is downward-closed in $\cat C$ and upward-closed in $\cat D$, a kind of
``hetero-ordering''.
%
Given profunctors $R, S : \cat C \pto \cat D$, a homomorphism from
$R$ to $S$ is simply a natural transformation, which in the preordered
setting is simply an implication of relations.

In addition to their usefulness for formalizing category theory, one
reason we take profunctors to be a basic concept of \vett{} is that they
allow us to give a \emph{universal property} for the type of ``morphisms
in a category ${\cat C}$''.  This is analogous to how the $J$ elimination
rule for the identity type in Martin-L\"of type theory gives a universal
property for morphisms in a groupoid (the special case of a category
where all morphisms are
invertible)~\cite{hofmannstreicher,awodey,voevodsky}. The reason
profunctors are useful for this purpose is that, for any category $\cat
C$, $\text{Hom}_{\cat C} : \cat C \pto \cat C$ is a profunctor. On
preorders this is just the preorder's ordering relation
itself. Moreover, the hom profunctor is the unit for a composition of
profunctors $R \odot S$ which is defined as a \emph{co-end}. The
heteromorphisms of $(R \odot S)(c,e)$ are triples $(d \in \cat D, f \in
R(c,d), g \in S(d,e))$ modulo the equivalence relation generated by
$(d,f,g) ~ (d', f', g')$ if there exists $h \in D(d,d')$ such that
$h\circ f = f'$ and $g'\circ h = g$. The composition of profunctors is a
generalization of the composition of relations, and just as the equality
relation is the identity for the composition of relations, the hom
profunctor is the identity for this composition.  The unit law for the
hom profunctor can be seen as a ``morphism induction'' principle,
analogous to the ``path induction'' used in homotopy type theory (though
in this paper we consider only ordinary 1-dimensional categories, not
higher generalizations).

%% orphaned --- move later? 
%% For any profunctor $R : \cat C \pto \cat D$ and functors $F : \cat C'
%% \to \cat C$ and $G : \cat D' \to \cat D$, by composition we can define a
%% profunctor $R(F,G)$, the ``restriction'' of $R$ along $F$ and $G$.

\paragraph{Paper Overview.}

\begin{enumerate}
\item In \cref{sec:syntax} we introduce the syntax of \vett{}.

\item In \cref{sec:examples} we demonstrate how to use our syntax for
  formal category theory.

\item In \cref{sec:semantics}, we develop some model theory for
  \vett{}, including a sound and complete notion of categorical model
  and sound interpretation in virtual equipments modeling ordinary,
  enriched and internal category theory.

\item In \cref{sec:discussion}, we discuss related type theories and
  potential extensions to model multi-variable and higher-order
  categorical constructions.
\end{enumerate}

\section{Syntax of VETT}

@@

Next, in Figure~\ref{fig:set-connectives}, we cover the connectives for
the sets/profunctors, which classify elements/natural transformations
(the $\beta/eta$-rules are in the appendix).
%% %
%% The top of the figure includes the formation rules for tensor products
%% $\tensorexistsXwithYandZ \beta P Q$, the covariant $\homrallXYtoZ
%% \alpha R P$ and contravariant homs $\homlallXYtoZ {\alpha} R P$ and
%% the unit type $\punitinXfromYtoZ {\cat C} {a_1}{a_2}$, and
%% the cartesian unit and product types $1$ and $R \times S$.  
%% %
%% The remainder of the figure gives the introduction and elimination rules
%% for these sets .  
%% %
First, the unit set is our syntax for the profunctor of morphisms of a
category $\cat C$, which is well-formed when given two objects of the
same category.  
%
Its introduction and elimination rules are analogous to the usual rules
for equality in Martin-L\"of type theory.  The introduction rule is the
identity morphism (reflexivity) and the elimination rule is an induction
principle: we can use a term of $s : \punitinXfromYtoZ {\cat C} A B$ by
specifying the behavior when $s$ is of the form $\punitrefl \alpha$ in
the form of a continuation $\alpha. t$.  Like the $J$ elimination rule
for equality in Martin-L\"of type theory, $P$ must be ``fully general'',
i.e. well-typed for variables $\alpha$ and $\beta$.  Those familiar with
linear logic~\cite{polakawpfenning} might expect a more general rule,
with the continuation $t$ typed in a context $\Phi_l \jnctx \Phi_r$ and
$\Phi_l \jnctx \Phi \jnctx \Phi_r$ as the context of the conclusion of
the rule.  Because of dependency, this is not necessarily well-formed in
cases where the endpoints $A$ and $B$ of $\punitinXfromYtoZ{\cat C} A B$
are not fully general (i.e. not distinct variables).  However, the
instances of this more general rule that do type check are derivable
from our more restricted rule using right/left-hom types.

@@

    \\
    \text{Tensor product:}\\
    \inferrule
    {\cat D \isSmall\\\\
      \Gamma \pipe d^-\Xi; \beta:D\vdash P \isaSet\\\\
     \Gamma \pipe \beta:D;d^+\Xi \vdash Q \isaSet}
    {\Gamma \pipe \Xi \vdash \tensorexistsXwithYandZ {\beta:D} P Q \isaSet}
    \quad
  \inferrule
  {\Gamma \pipe d^+\Psi_s \vdash b : \cat D\\\\
   \Gamma \pipe \Psi_s \vdash s : P[b/\beta] \\\\
   \Gamma \pipe \Psi_t \vdash t : Q[b/\beta]}
  {\Gamma\pipe\Psi_s \jnctx \Psi_t\vdash \tensorintroatXwithYandZ b s t : \tensorexistsXwithYandZ {\beta : \cat D} P Q}
  \quad
  \inferrule
  {
   \Gamma\pipe \Phi_m \jnctx x:P, \beta:\cat D, y:Q \jnctx \Phi_r \vdash t : R\\\\
   \Gamma \pipe\Phi_m \vdash s : \tensorexistsXwithYandZ {\beta} P Q}
  {\Gamma\pipe\Phi_l\jnctx \Phi_m \jnctx \Phi_r \vdash \tensorelimWkontZ s {x.\beta.y. t} : R}
  \\
  %% \tensorintroatXwithYandZ B M N [\phi,B'/\beta',\psi] = \tensorintroatXwithYandZ {B[B'\beta]} {M[\phi]} {N[\psi]}

@@

\section{Formal Category Theory in VETT}

@@

\section{Semantics}

@@

Our use of dependent type theory as a metalanguage, with new judgements
for the virtual equipment structure dependent on the meta-language
context, is similar to the indexed type theories developed by
\citet{isaev}.

@@

>>>>

\begin{abstract}
  We present a domain-specific type theory for constructions and proofs
  in category theory. The type theory axiomatizes notions of category,
  functor, profunctor and a generalized form of natural
  transformations. The type theory imposes an ordered linear restriction
  on standard predicate logic, which guarantees that all functions
  between categories are functorial, all relations are profunctorial,
  and all transformations are natural by construction, with no separate
  proofs necessary. Important category theoretic proofs such as the
  Yoneda lemma and Co-yoneda lemma become simple type theoretic proofs
  about the relationship between unit, tensor and (ordered) function
  types, and can be seen to be ordered refinements of theorems in
  predicate logic.  The type theory is sound and complete for a
  categorical model in \emph{virtual equipments}, which model both
  internal and enriched category theory. While the proofs in our type
  theory look like standard set-based arguments, the syntactic
  discipline ensure that all proofs and constructions carry over to
  enriched and internal settings as well.
\end{abstract}

@@

Category theory is a branch of mathematics that studies
higher-dimensional typed algebraic structures.
%
Originally developed for applications to homological algebra, it was
quickly discovered that categorical structures were common in logic and
computer science.  Formal systems like logics, type theories and
programming languages typically have sound and complete models given by
notions of structured categories,\cite{lawvere,goguen,lambek,moggi}.
%
This Curry-Howard-Lambek correspondence~\cite{curry-howard-lambek}
applies to simply typed lambda calculus~\cite{lambekscott},
computational lambda calculus~\cite{moggi}, linear logic~\cite{girard}
dependent type theory\cite{cartmell1986}, and many other type theories
designed based on category-theoretic semantics.
%
The syntax of a type theory should present an initial object in its
category of models, a category-theoretic reformulation of logical
soundness and completeness.

% TODO: do we have citations for this?
While this research program has been quite successful,
category-theoretic techniques in computer science are sometimes
criticized for being overly complex and hard to learn, especially for
computer scientists who do not have a strong background in pure
mathemantics.
%
While in some ways this could be remedied by better introductory
materials, in some ways category theory is objectively complex.
%
In a traditional set-theoretic formulation, notions such as adjoint
functors and limits produce a proliferation of ``naturality'' and
``functoriality'' side-conditions that must be discharged.
%
For example, when constructing an adjoint pair of
functors between two categories, a na\"ive approach would define all of
the data of the action on objects, action on arrows, prove the
functoriality of such actions, as well as construct two families of
transformations, prove they are natural and then finally proving a pair
of equalities relating compositions of natural transformations.
%
Carrying out these proofs explicitly is quite tedious and many newcomers
to are left with the impression that category theory is full of long,
but ultimately trivial constructions.
%
This complexity is compounded when we move from ordinary category theory to
enriched and internal category theory, where constructions must be
additionally proven continuous, monotone, etc, in addition to natural
or functorial.
%
Moreover, these generalizations are often what is needed for programming
language applications; for example, domain- and metric- and
step-index-enriched categories have been used to model recursive
programming languages and internal categories have been used to model
parametricity and gradual
typing\cite{order-enriched-cats,topos-of-trees,reflexive-graphs,double-cats-gradual-typing}.

Fortunately, the tools of category theory itself can be employed to
simplify this complexity, specifically the tools of \emph{higher}
category theory.
%
As an analogy in differential calculus, when an adept analyst writes
down a function, they do not expand out the $\epsilon\mathord{-}\delta$
definition of continuity for a function and proceed from first
principles, but rather use certain \emph{syntactic principles} for
defining functions that are continuous by construction --- e.g. that
composition of continuous functions is continuous and that previously
defined functions are known to be continuous.
%
Similar principles apply to category theory itself: functors and
natural transformations are closed under composition and whiskering
operations, and experienced category theorists rely on these syntactic
principles to eliminate the tedium of explicit proofs.
%
In the case of category theory, these principles can be formalized
using algebraic structures such as 2-categories, bicategories,
(virtual) double categories and pro-arrow equipments\cite{TODO}, an
approach known as \emph{formal category theory}.
%
In these structures, rather than defining notions of category, functor
and natural transformation from first principles, they are axiomatized
in a manner similar to how a category axiomatizes a notion of space
and homomorphism.
%
The formal approaches also apply to enriched and internal settings, so
\emph{formal} category theory proofs apply to these settings.
%
A downside is that these algebraic structures are quite complicated, and
practitioners typically employ either an algebraic combinator syntax or
a 2-dimensional diagrammatic language that can be quite beautiful and
elegant, but is also somewhat removed from the traditional formulation
of category theory in terms of sets and functions.

In this work, we apply the techniques of categorical logic to define a
more familiar logical syntax for carrying out constructions and proofs
in formal category theory.  
%
We call the resulting theory \emph{virtual equipment type theory}
(VETT) as (hyperdoctrines of) virtual equipments provide a sound and
complete notion of model for the theory.
%
VETT provides syntax for categories, functors, profunctors, and natural
transformations, which are defined using familiar term syntax and
$\beta\eta$ reasoning principles for $\lambda$-functions, bound
variables, tuples, etc.  
%
By adhering to a \emph{syntactic discipline}, the logic guarantees
that all functor terms are automatically functorial, all and natural
transformation terms are natural.
%
More specifically, the syntax for transformations is a kind of
\emph{indexed, ordered linear} $lambda$ calculus, where the indexing
ensures that transformations are correctly natural and the ordering and
linearity ensure that the proofs are valid in a large class of enriched
and internal categories, such as enrichment in a non-symmetric monoidal
category.
%
VETT provides an alternative alternative to algebraic and
string-diagram syntaxes for working with virtual equipments,
similar to how the lambda calculus provides an alternative to
categorical combinators and string diagram calculi for cartesian
closed categories.

%% The type theoretic syntax allows us to write all of our 
%% constructions are written as if they were
%% ordinary functors and profunctors in set-based category theory.

%%
%% DRL: I know what you mean, but I think conflating set theory and
%% type theory will put some people off... 
%% 
%% This is analogous to the situation in intuitionistic type theory,
%% where the syntax is similar to that of informal set theory, but by
%% restricting to intuitionistic reasoning, the constructions can all be
%% interpreted in models where types are spaces and terms are continuous
%% functions ``for free''.
%

The syntax of \vett{} is an indexed, ordered linear, proof-relevant
variant of predicate logic over a unary type theory.
%
Just as a predicate logic has a notion of type, term, relation and
implication, \vett{} is based on four analogous category theoretic
concepts: categories, functors, profunctors and natural transformations
of profunctors.
%
Categories are treated like types, and the unary functors we consider in
this paper are each represented by a term whose type is a category and
whose one free variable ranges over a category.
%
The analogue of a relation is a \emph{profunctor} (defined below), which
are written as a kind of set with free category variables.  Like the
restriction to unary functors, we restrict to profunctors with two free
variables.
%
The logic is proof-relevant in that the implications of relations are
generalized to natural transformations of profunctors, and we use a
$\lambda$-calculus notation to describe these ``proof terms''.
%
We can strenghten this analogy into a formal construction: any
construction in \vett{} can be erased to a corresponding construction
or proof in predicate logic, as sets, functions and relations define a 
(somewhat degenerate) virtual equipment.
%% %
%% Our syntax highlights this analogy in that some of our constructions
%% are generalizations of set comprehension or power-sets and we use
%% similar notation.

While the restricted syntax developed in this paper does not express
important concepts such as functor categories or opposite categories,
the restriction is natural in that it corresponds exactly to a
well-understood notion of model, virtual equipments, which can express a
great deal of fundamental results and constructions in category
theory~\cite{elements-of-oo-category-theory,
  enriched-indexed-categories}.
%
Moveover, we can work around these unary/binary restrictions to some
extent by viewing the type theory as a domain-specific language embedded
in a metalanguage.  For example, while we cannot talk about functor
categories, we can state a theorem that quantifies over functors using
the meta-language's ``external'' universal quantifier (which does not
have automatic functoriality/naturality properties).  To support this,
\vett includes a third layer, an extensional dependent type theory in
the style of Martin-L\"of type theory. All of our ordered predicate
logic judgments are also be indexed by a context from this dependent
type theory, and the type theory includes universe types for categories,
functors, profunctors and natural transformations.  This allow us to
formalize theorems the logic is too restrictive to encode, analogous to
2-level~\cite{voevodsky12,altenkirchXX} or indexed type
theories~\cite{isaev,linearlf,vakar,krishnaswamibenton}.

@@

While we emphasize the applications to enriched and internal category
theory in this work, there is potential for more direct application to
programming language semantics.
%
Ordinary predicate logic is the foundation for logical presentations of
logical relations, such as Abadi-Plotkin logic for parametricity and
LSLR and Iris for step-indexed logical relations proofs
\cite{abadi-plotkin,lslr,iris}.
%
We conjecture that \vett{} might similarly serve as the foundation for
a logic of \emph{ordered} structures, which abound in applications:
rewriting and approximation relations can both be modeled as orderings
and logical relations involving these structures are proven to respect
orderings: operational logical relations must be downward-closed and
approximation relations should satisfy transitivity.
%
Just as LSLR and Iris release the user from the syntactic burden of
explicit step-indexing, \vett{} may be used to release the user from
the syntactic burden of proving downward-closure or transitivity
side-conditions.
%
Additionally, \vett{} may serve as the basis of a future domain specific
proof assistant for category theoretic proofs.  To pilot-test this, we
have formalized the syntax of \vett{} in Agda 2.6.2.2, using the rewrite
mechanism to make \vett's substitution and $\beta$-reduction rules
definitional
equalities.\footnote{\url{https://github.com/maxsnew/virtual-equipments/blob/master/agda/STC.agda}}
We have used this lightweight implementation to check a number of
examples.

\paragraph{Basics of Profunctors.}
While we assume the reader has some background knowledge of category
theory, we briefly define profunctors, which are not included in many
introductory texts.  Recall that a category $\cat C$ has a collection of
objects and morphisms with identity and composition, and a functor $F :
\cat C \to \cat D$ is a function on objects and a function on morphisms
that preserves identity and composition.  A category can be thought of
as a generalization of a preordered set, which has a set of elements and
a binary \emph{relation} on its objects satisfying reflexivity and
transitivity.  A category is then a \emph{proof-relevant preorder},
where morphisms are the proofs of ordering, and the reflexivity and
transitivity proofs must identity and unit equations. A functor is then
a \emph{proof-relevant monotone function}.  Given categories $\mathcal
C$ and $\mathcal D$, a profunctor $R$ from $\mathcal C$ to $\mathcal D$,
written $R : \cat C \pto \cat D$ is a functor $R : {\cat C}^{op} \times
{\cat D} \to \Set$.  Because a profunctor outputs a $\Set$ rather than a
proposition, it is itself a \emph{proof-relevant relation}.  Thinking of
categories as proof-relevant preorders, functoriality says that the
profunctor is downward-closed in $\cat C$ and upward-closed in $\cat D$.
Given profunctors $R, S : \cat C \pto \cat D$, a homomorphism from
$R$ to $S$ is simply a natural transformation, which in the preordered
setting is simply an implication of relations.

%% We can think of each set $R(c,d)$ as a set of
%% ``heteromorphisms'' from $c$ (an object of $\cat C$) to $d$ (an object
%% of $\cat D$).  These ``heteromorphisms'' differ from the usual
%% ``homomorphisms'' in a single category $\cat C$ because they relate two
%% objects from two different categories.  Functoriality then defines a
%% notion of composition of heteromorphisms with homomorphisms in $\cat C$
%% and $\cat D$.

Profunctors are very useful for formalizing category theory, but an
additional reason we make them a basic concept of \vett{} is that they
allow us to give a \emph{universal property} for the type of ``morphisms
in a category ${\cat C}$''.  This is analogous to how the $J$
elimination rule for the identity type in Martin-L\"of type theory gives
a universal property for morphisms in a groupoid (the special case of a
category where all morphisms are
invertible)~\cite{hofmannstreicher,awodeywarren,voevodsky}. The reason
profunctors are useful for this purpose is that, for any category $\cat
C$, $\text{Hom}_{\cat C} : \cat C \pto \cat C$ is a profunctor. On
preorders this is just the preorder's ordering relation
itself. Moreover, the hom profunctor is the unit for a composition of
profunctors $R \odot S$ which is defined as a \emph{co-end}.
%% The
%% heteromorphisms of $(R \odot S)(c,e)$ are triples $(d \in \cat D, f \in
%% R(c,d), g \in S(d,e))$ modulo the equivalence relation generated by
%% $(d,f,g) ~ (d', f', g')$ if there exists $h \in D(d,d')$ such that
%% $h\circ f = f'$ and $g'\circ h = g$.
The composition of profunctors is a generalization of the composition of
relations, and just as the equality relation is the identity for the
composition of relations, the hom profunctor is the identity for this
composition.  The unit law for the hom profunctor can be seen as a
``morphism induction'' principle, analogous to the ``path induction''
used in homotopy type theory (though in this paper we consider only
ordinary 1-dimensional categories, not higher generalizations).

%% orphaned --- move later? 
%% For any profunctor $R : \cat C \pto \cat D$ and functors $F : \cat C'
%% \to \cat C$ and $G : \cat D' \to \cat D$, by composition we can define a
%% profunctor $R(F,G)$, the ``restriction'' of $R$ along $F$ and $G$.
The remainder of this paper is organized as follows.  
In \cref{sec:syntax} we introduce the syntax of \vett{}.  In
\cref{sec:examples} we demonstrate how to use our syntax for formal
category theory.  In \cref{sec:semantics}, we develop some model theory
for \vett{}, including a sound and complete notion of categorical model
and sound interpretation in virtual equipments modeling ordinary,
enriched and internal category theory.  In \cref{sec:discussion}, we
discuss related type theories and potential extensions.

\section{Syntax of VETT}
\label{sec:syntax}

@@

Next, in Figure~\ref{fig:set-connectives}, we cover the connectives for
the sets/profunctors, which classify elements/natural transformations
(the $\beta/eta$-rules are in the appendix).
%% %
%% The top of the figure includes the formation rules for tensor products
%% $\tensorexistsXwithYandZ \beta P Q$, the covariant $\homrallXYtoZ
%% \alpha R P$ and contravariant homs $\homlallXYtoZ {\alpha} R P$ and
%% the unit type $\punitinXfromYtoZ {\cat C} {a_1}{a_2}$, and
%% the cartesian unit and product types $1$ and $R \times S$.  
%% %
%% The remainder of the figure gives the introduction and elimination rules
%% for these sets .  
%% %
First, the unit set is our syntax for the profunctor of morphisms of a
category $\cat C$, which is well-formed when given two objects of the
same category.  
%
Its introduction and elimination rules are analogous to the usual rules
for equality in Martin-L\"of type theory.  The introduction rule is the
identity morphism (reflexivity) and the elimination rule is an induction
principle: we can use a term of $s : \punitinXfromYtoZ {\cat C} A B$ by
specifying the behavior when $s$ is of the form $\punitrefl \alpha$ in
the form of a continuation $\alpha. t$.  Like the $J$ elimination rule
for equality in Martin-L\"of type theory, $P$ must be ``fully general'',
i.e. well-typed for variables $\alpha$ and $\beta$; this is because for
distinct variables $\alpha$ and $\beta$, $\punitinXfromYtoZ{\cat
  C}{\alpha}{\beta}$ denotes the unit, which has a universal property,
but $\punitinXfromYtoZ{\cat C} A B$ denotes a restriction of a unit,
which in general does not.  Those familiar with linear
logic~\cite{polakawpfenning} might expect a more general rule, with the
continuation $t$ typed in a context $\Phi_l \jnctx \Phi_r$ and $\Phi_l
\jnctx \Phi \jnctx \Phi_r$ as the context of the conclusion of the rule.
Because of dependency, this is not necessarily well-formed in cases
where the endpoints $A$ and $B$ of $\punitinXfromYtoZ{\cat C} A B$ are
not fully general (i.e. not distinct variables).  However, the instances
of this more general rule that do type check are derivable from our more
restricted rule using right/left-hom types.

@@

    \\
    \text{Tensor product:}\\
    \inferrule
    {\cat D \isSmall\\\\
      \Gamma \pipe d^-\Xi; \beta:D\vdash P \isaSet\\\\
     \Gamma \pipe \beta:D;d^+\Xi \vdash Q \isaSet}
    {\Gamma \pipe \Xi \vdash \tensorexistsXwithYandZ {\beta:D} P Q \isaSet}
    \quad
  \inferrule
  {\Gamma \pipe d^+\Psi_s \vdash b : \cat D\\\\
   \Gamma \pipe \Psi_s \vdash s : P[b/\beta] \\\\
   \Gamma \pipe \Psi_t \vdash t : Q[b/\beta]}
  {\Gamma\pipe\Psi_s \jnctx \Psi_t\vdash \tensorintroatXwithYandZ b s t : \tensorexistsXwithYandZ {\beta : \cat D} P Q}
  \quad
  \inferrule
  {
   \Gamma\pipe \Phi_l \jnctx x:P, \beta:\cat D, y:Q \jnctx \Phi_r \vdash t : R\\\\
   \Gamma \pipe\Phi_m \vdash s : \tensorexistsXwithYandZ {\beta} P Q}
  {\Gamma\pipe\Phi_l\jnctx \Phi_m \jnctx \Phi_r \vdash \tensorelimWkontZ s {x.\beta.y. t} : R}
  \\
  %% \tensorintroatXwithYandZ B M N [\phi,B'/\beta',\psi] = \tensorintroatXwithYandZ {B[B'\beta]} {M[\phi]} {N[\psi]}

@@

\section{Formal Category Theory in VETT}
\label{sec:examples}

@@

\section{Semantics}
\label{sec:semantics}

@@

@@

%% \subsubsection*{Acknowledgements}
%% Grants stuff...? The authors would like to thank David Jaz Myers, Emily Riehl, Mike
%% Shulman, Dominic Verity for helpful feedback on this work.
